---
title: "Problem Set 3"
author: "Bailey Kretzler & Sarah Morris"
date: "10/5/2020"
output: html_document
---

In teams of two, address the following questions:

What species (and location if available) did you choose? Why did you choose the species?
What question do you want to answer about this population (e.g. population status, best management strategies)?
Calculate eigenvalue, stable age distribution, elasticity, and sensitivity. What does this tell you about the population?
Using the calculations in part (c), or additional calculations, address the question you proposed in part (b).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**(a)**

We selected the species *Phaseoulus lunatus*, commonly known as lima bean, in part because it is a food crop but also because of interesting features in the matrix. The matrix describes a threatened wild population of lima bean in Costa Rica. This population is a valuable genetic resource for biodiversity conservation and varietal improvement in other beans. 

**(b)** 

Within the matrix, it is clear that lima beans have strong seed dormancy. Therefore, it would be interesting to follow the trajectory of the associated paper and examine how seed dormancy impacts survival and growth. Specifically, if we decrease seed dormancy and increase seed germination, what happens to species survival in terms of the eigen value? These changes would be reflected in the matrix by increasing the rate at which individuals move from stage G~0~ to L~1~ and proportionally decreasing the number of individuals moving from G~0~ to G~1~. 


**c** To calculate summary statistics about the population, first we need to load it into R and visualize it:
```{r creating/finding matrix}

#alt way to load matrix
grep(pattern = 'Phaseolus', compadre$metadata$Genus)
compadre$metadata$Genus[5735]
compadre$mat[5735][[1]]$matA

Phasmat <- compadre$mat[5735][[1]]$matA

# creating the matrix
#data<- c(0,0.16,0.12,0.08,0,0,0,0,0.14,0.09,0,0,0,0,0,0.01,0,0,0.42,0,0,0.01,0.05,0,28.9,0,0,0,0.16,0.18,104,0,0,0,0,0.25) #data in matrix by column

#Phasmat<- matrix(data = data, nrow = 6, ncol =  6) #creating matrix and inputting above data

colnames(Phasmat) = c("G0", "G1", "J", "L1", "L2", "L3") #assigning the life stage IDs to the column names

Phasmat #viewing matrix

L <- Phasmat # reassigning matrix to object L for leslie matrix

```

In lieu of extracting the matrix from the website, we created the matrix here.

```{r projecting the population behavior}
require(tidyr)
t = 30 # setting time = to 30
Nt = matrix(nrow = nrow(L), ncol = t+1 , data = 0) # creating an initial population matrix


Nt[,1] = 50 #where all life stages are at size 50 in year 1
Nt #checking Nt

L # as defined above

#for loop to determine the size of each life stage across 50 years
for (i in 1:t)
{
	Nt[, i + 1] <- L %*% Nt[,i]
}

Pop <- t(Nt)

time =c(1:31)
Pop_time <- as.data.frame(cbind(Pop, time))
colnames(Pop_time) = c("G0", "G1", "J", "L1", "L2", "L3", "time")
Pop_time <- gather(Pop_time, key = "time", value = "pop")
repeat_t =c(c(1:31),c(1:31),c(1:31),c(1:31),c(1:31),c(1:31))
Pop_time <- cbind(Pop_time, repeat_t)

require(ggplot2)

ggplot(data = Pop_time, aes(x = repeat_t, y = pop, colour = time)) + geom_point(size = 3) + theme_bw() + geom_line() + ggtitle("Population Projection for Original Leslie Matrix", subtitle = "Phaseolus lunatus") +labs(x = "Time in Years", y = "Population size" ) +scale_color_discrete(name = "Life Stage")

```
Based on the above population projection it seems that at the present fecundity, survival, and growth rates all Lima bean life stages will decrease with time. G0 will see a slight increase due to high fecundity rates but will continue to drop as the number of reproductive lignified individuals (L1, L2, L3) declines.

So why is this the case? What do the eigenvalues and vectors look like?

```{r finding eigenvalues and vectors}

require(popbio)

#our numbers are slightly different from paper - why?
eigen1 = eigen.analysis(L)

eigen1$lambda1 #eigenvalue
eigen1$stable.stage #stable stage distribution
eigen1$sensitivities #sensitivity matrix
eigen1$elasticities #elasticity matrix

eigen1$sensitivities[4,1] #sensitivity of G0 to L1
eigen1$sensitivities[1,4:6] #sensitivity for fecundity values of L1 to L3
eigen1$sensitivities[2,1] #sensitivity of G0 to G1

```
The eigen value of 0.7675902 indicates that the population is in exponential decay (because it's less than 1). The right eigenvector, or stable stage distribution which is proportional here, indicates that at a stable state the majority of individuals will exist in stage G0, followed by G1, and J. The least number of individuals will be in the lignified stages. It is possible this is a result of high fecundity. The sensitivity matrix indicates that, given we are only looking at realistic examples, that increasing the rate of individuals moving from G0 to L1 will have the greatest impact on the eigenvalue. Specifically, for every unit increase in G0 to L1, the eigenvalue will change by 1.83299099. The rate of growing from G0 to G1 has some impact on the eigenvalue but much less, with the rate of growing from G0 to J being even lower. Additionally, changes in fecundity rates of life stages L1 - L3 will have a much lower impact on the eigenvalue. The elasticity matrix indicates the movement from G0 to L1 has the greatest proportional impact on the eigenvalue. Specifically every unit change in G0 to L1, the eigen value changes by 19%. (??????? is this correct?)



From here we would like to know how the stable age distribution, eigenvalues, eigenvectors and sensitivity/elasticity changes if we increase the rate of individuals moving from G0 to L1 and decrease the number of individuals moving from G0 to G1. To do this we will lower the G0 - G1 value by 0.08 and raise the G0 - L1 value by 0.08 and subsequently reexamine the populations behavior.

```{r creating new matrix}


Phasmat2 <- compadre$mat[5735][[1]]$matA

Phasmat2[2,1] = 0.08 #changing G0 to G1 value to 0.08
Phasmat2[4,1] = 0.16 #changing G0 to L1 value to 0.16

colnames(Phasmat2) = c("G0", "G1", "J", "L1", "L2", "L3") #assigning the life stage IDs to the column names

Phasmat2 #viewing matrix

L2 <- Phasmat2 # reassigning matrix to object L for leslie matrix

```

Above we redefine the matrix and assign it a new name. Now let's examine how the population looks over time.

```{r projecting the new population behavior}
t = 30 # setting time = to 30
Nt2 = matrix(nrow = nrow(L2), ncol = t+1 , data = 0) # creating an initial population matrix


Nt2[,1] = 50 #where all life stages are at size 50 in year 1
Nt2 #checking Nt

L2 # as defined above

#for loop to determine the size of each life stage across 50 years
for (i in 1:t)
{
	Nt2[, i + 1] <- L2 %*% Nt2[,i]
}

Pop2 <- t(Nt2)

time =c(1:31)
Pop_time2 <- as.data.frame(cbind(Pop2, time))
colnames(Pop_time2) = c("G0", "G1", "J", "L1", "L2", "L3", "time")
Pop_time2 <- gather(Pop_time2, key = "time", value = "pop")
repeat_t2 =c(c(1:31),c(1:31),c(1:31),c(1:31),c(1:31),c(1:31))
Pop_time2 <- cbind(Pop_time2, repeat_t2)


ggplot(data = Pop_time2, aes(x = repeat_t2, y = pop, colour = time)) + geom_point(size = 3) + theme_bw() + geom_line() + ggtitle("Population Projection for Adjusted Leslie Matrix", subtitle = "Phaseolus lunatus") +labs(x = "Time in Years", y = "Population size" ) +scale_color_discrete(name = "Life Stage")

```
It seems that the population is still in exponential decline but at a much slower rate than before. How is this reflected in the parameters?

```{r calculating new eigenvectors/values}

eigen.analysis(L2) 


```
How does this look if we take this further and decrease the value of G0 to G1 to 0.01 and the value of G0 to L1 to 0.23? We can use iteration to examine how the relative change in these rates impacts the eigenvalue.

```{r projecting the new population behavior}

G0L1 = seq(from =  0.08 , to = 0.23, by = 0.01)
G0G1 = seq(from = 0.16, to = 0.01, by = -0.01)
length(G0L1) == length(G0G1)
eigenmat = matrix(data = 0, nrow = length(G0L1), ncol = 1)# creating an initial population matrix

for (i in 1:length(G0L1)) { # for length of sequence
  L[2,1] = G0G1[i] #make G0to G1 in L matrix equal to the i position in the corresponding vector
  L[4,1] = G0L1[i] #make G0to L1 in L matrix equal to the i position in the corresponding vector
  eigenmat[i,] = eigen.analysis(L)$lambda1 #extract eigenvalues
}
print(eigenmat)

G1_L1 = G0L1/G0G1
eigendf = as.data.frame(eigenmat) # make data frame
eigendf = cbind(eigendf, G1_L1) # bind corresponding values

ggplot(data = eigendf, aes(x = G1_L1, y = V1)) + geom_point(size = 3) + theme_bw() + geom_line() + ggtitle("Imapact of proportional change of G0 to L1 and G0 to G1 \ngrowth rates on Eigenvalue", subtitle = "Phaseolus lunatus") +labs(x = "Ratio of Growth Rates (G0L1/G0G1)", y = "Eigenvalue" )



```

Here it is clear that as the ratio of the G0L1 growth rate to the G0G1 growth rate increases, so does the eigenvalue. The eigenvalue begins to approach 1 and the rate of exponential decay slows. While this would prolong the species survival, it does not prevent populations from dropping toward 0 (extinction). The growth rate of G1 to L1 would have to increase even further. This increase could be the result of decreased death rate instead on decreased growth from G0 to G1.

**(d)**
